Effects of bariatric surgery and weight loss on resting state functional connectivity

### 2021-01-07 manuscript/draft_rsfmri.Rmd
fix figure design

## Functional preprocessing
### Minimal functional preprocessing
The function preprocessing module included the removal of first four volumes, motion correction (FSL's `MCFLIRT`), fieldmap distortion correction (FSL's `fsl_prepare_fieldmap` and `FUGUE`) and coregistration to the subject’s individual longitudinal anatomical space (FreeSurfer's `bbregister`). In more detail, the transformations derived from the latter three steps were combined into one and applied in a single step.
For further analysis and ICA-AROMA processing, the minimally preprocessed data were intensity normalized and smoothed with a 6mm Gaussian kernel (`fslmaths -kernel gauss 2.548`).
*This part of the analysis was described in the preregistration as "After removing the first four volumes, transforms for motion correction (FSL’s MCFLIRT), fieldmap distortion correction (based on fieldmap and FSL’s FUGUE) and coregistration to the subject’s individual anatomical space (using FREESURFER’s bbregister) are calculated."*

@@

<!-- R analysis of aggregated values -->
We analyzed the aggregate connectivity values using linear mixed models in the R package lme4 [@Bates_2007].
Following the preregistration, we first investigated a time-by-group interaction between baseline and follow-up.
We used a model comparison of R1 and R0 models with `anova`:
`R1= lmer(agg FC ~ timepoint*group + (1|subj))`
`R0 = lmer(agg FC ~ timepoint + group + (1|subj))`
We performed a sensitivity analysis adjusting for baseline age, sex, average of mean FD of both time points and baseline BMI. Again, the time-by-group point interaction was considered significant if a model comparison between R1 and Null models using anova-function yielded a p<0.05.
Finally, we estimated the above-mentioned interaction model for all three time points. Again, the interaction was considered significant if a model comparison between R1 and R0 using `anova` had p<0.05.

@@

for (i in 1:nrow(final)) {
  tmp = strsplit(toString(final[i, "scan_dir"]), '/')[[1]][8]
  conf = read.csv(
    paste0("/data/pt_02161/preprocessed/resting/detailedQA/metrics/",
      tmp, "/confounds.csv"))
  tmp = cor.test(conf$DVARS, conf$FD)
  final[i, "corr_DVARS_FD"] = tmp$estimate
  final[i, "mean_DVARS"] = mean(conf$DVARS, na.rm = TRUE)
  final[i, "max_DVARS"] = max(conf$DVARS, na.rm = TRUE)
  final
}

@@

FigureList <- figDvarsmFD(final)
(FigureList$fig1 | FigureList$fig2) / FigureList$fig3

@@

### Summary
Utilizing resting-state functional MRI on morbidly obese individuals that either underwent or awaited bariatric surgery over the course of one year.
<!-- general --> Building on the evidence for post-surgery changes in functional activity, this study explored the longitudinal relationship between surgery-induced weight loss and functional integrity in target networks indicated by resting-state functional connectivity. Over the course of one year we collected neuroimaging and anthropometric data from `r cs["Sum","IG"]` morbidly obese patients undergoing bariatric surgery and compared them to the longitudinal data of  `r cs["Sum","KG"]` patients on the waiting list.
<!-- model specific --> Looking at time-by-group interactions, we primarily investigated whether changes in functional connectivity over time vary depending whether participants have received surgery or not, while controlling for individual variance in resting-state connectivity over time and at baseline. Beyond that, also main effects of the intervention and of the time were examined.
The second model further investigates if changes in connectivity are a function of an average BMI (controlled for individual variance in BMI) and the longitudinal change in BMI.
<!-- sensitivity analyses -->To get a better understanding of our results, we conducted multiverse exploratory analyses representing different choices regarding preprocessing pipelines, inclusion of covariates and subsets of the data relevant for interpretation.

### Support of original hypotheses
<!-- false positive -->
To summarize, while our primary hypothesis on group time interactions could not be confirmed, there was a significant relationship between functional activation 
and the continuous BMI measures for some but not all analyses.

>>>>

## Functional preprocessing

### Minimal functional preprocessing
The function preprocessing module included the removal of first four volumes, motion correction (FSL's `MCFLIRT`), fieldmap distortion correction (FSL's `fsl_prepare_fieldmap` and `FUGUE`) and coregistration to the subject’s individual longitudinal anatomical space (FreeSurfer's `bbregister`). In more detail, the transformations derived from the latter three steps were combined into one and applied in a single step.
For further analysis and ICA-AROMA processing, the minimally preprocessed data were intensity normalized and smoothed with a 6mm Gaussian kernel (`fslmaths -kernel gauss 2.548`).
*This part of the analysis was described in the preregistration as "After removing the first four volumes, transforms for motion correction (FSL’s MCFLIRT), fieldmap distortion correction (based on fieldmap and FSL’s FUGUE) and coregistration to the subject’s individual anatomical space (using FREESURFER’s bbregister) are calculated."*

@@

<!-- R analysis of aggregated values -->
We analyzed the aggregate connectivity values using linear mixed models in the R package lme4 [@Bates_2007].
Following the preregistration, we first investigated a time-by-group interaction between baseline and follow-up.
We used a model comparison of R1 and R0 models with `anova`:
`R1= lmer(agg FC ~ timepoint*group + (1|subj))`
`R0 = lmer(agg FC ~ timepoint + group + (1|subj))`
We performed a sensitivity analysis adjusting for baseline age, sex, average of mean FD of both time points and baseline BMI. Again, the time-by-group point interaction was considered significant if a model comparison between R1 and Null models using `anova` function yielded a $p < 0.05$.
Finally, we estimated the above-mentioned interaction model for all three time points. Again, the interaction was considered significant if a model comparison between R1 and R0 using `anova` had $p < 0.05$.

@@

for (i in 1:nrow(final)) {
  tmp = strsplit(toString(final[i, "scan_dir"]), '/')[[1]][8]
  conf = read.csv(
    paste0("/data/pt_02161/preprocessed/resting/detailedQA/metrics/",
      tmp, "/confounds.csv"))
  tmp = cor.test(conf$DVARS, conf$FD)
  final[i, "corr_DVARS_FD"] = tmp$estimate
  final[i, "mean_DVARS"] = mean(conf$DVARS, na.rm = TRUE)
  final[i, "max_DVARS"] = max(conf$DVARS, na.rm = TRUE)
  final
}
```

@@

FigureList <- figDvarsmFD(final)
(FigureList$fig1 | FigureList$fig2) / FigureList$fig3
```

@@

#### Summary
Utilizing resting-state functional MRI on morbidly obese individuals that either underwent or awaited bariatric surgery over the course of one year.
<!-- general --> Building on the evidence for post-surgery changes in functional activity, this study explored the longitudinal relationship between surgery-induced weight loss and functional integrity in target networks indicated by resting-state functional connectivity. Over the course of one year we collected neuroimaging and anthropometric data from `r cs["Sum","IG"]` morbidly obese patients undergoing bariatric surgery and compared them to the longitudinal data of  `r cs["Sum","KG"]` patients on the waiting list.
<!-- model specific --> Looking at time-by-group interactions, we primarily investigated whether changes in functional connectivity over time vary depending whether participants have received surgery or not, while controlling for individual variance in resting-state connectivity over time and at baseline. Beyond that, main effects of the intervention and of the time were examined.
The second model further investigates if changes in connectivity are a function of an average BMI (controlled for individual variance in BMI) and the longitudinal change in BMI.
<!-- sensitivity analyses -->To get a better understanding of our results, we conducted multiverse exploratory analyses representing different choices regarding preprocessing pipelines, inclusion of covariates and subsets of the data relevant for interpretation.

#### Support of original hypotheses
<!-- false positive -->
To summarize, while our primary hypothesis on group time interactions could not be confirmed, there was a significant relationship between functional activation 
and the continuous BMI measures for some but not all analyses.
